Modal type theory based on the intuitionistic epistemic logic

### 2018-01-14 article.tex
rewrite

@@

  \begin{center}
  \begin{prooftree}
  \AxiomC{$ $}
  \RightLabel{\scriptsize{ax}}
  \UnaryInfC{$\Gamma , x : A \vdash x : A$}
  \end{prooftree}
  \end{center}

  \begin{minipage}{0.5\textwidth}
    \begin{flushleft}
  \begin{prooftree}
  \AxiomC{$\Gamma, x : A \vdash M : B$}
  \RightLabel{$\rightarrow_i$}
  \UnaryInfC{$\Gamma \vdash \lambda x. M : A \to B$}
  \end{prooftree}

  \begin{prooftree}
  \AxiomC{ $\Gamma \vdash x : A$ }
  \AxiomC{ $\Gamma \vdash y : B$ }
  \RightLabel{$\times_i$}
  \BinaryInfC{$\Gamma \vdash \langle x, y \rangle : A \times B$}
  \end{prooftree}

  \begin{prooftree}
    \AxiomC{$\Gamma \vdash x : A$}
    \RightLabel{$\textbf{K}_I$}
    \UnaryInfC{$\Gamma \vdash {\bf pure \: } \: x : \textbf{K}A $}
  \end{prooftree}
    \end{flushleft}
  \end{minipage}

  \begin{minipage}{0.5\textwidth}
    \begin{flushright}
  \begin{prooftree}
  \AxiomC{$\Gamma \vdash f : A \to B$}
  \AxiomC{$\Gamma \vdash x : A$}
  \RightLabel{$\rightarrow_e$}
  \BinaryInfC{$\Gamma \vdash fx : B$}
  \end{prooftree}

  \begin{prooftree}
  \AxiomC{ $\Gamma \vdash M : A_1 \times A_2$ }
  \RightLabel{$\times_e$, $i \in \{ 1, 2 \}$}
  \UnaryInfC{$\Gamma \vdash \pi_i M : A_i$}
  \end{prooftree}

  \begin{prooftree}
    \AxiomC{$\Gamma \vdash \vec{M} : {\bf K} \vec{A}$}
    \AxiomC{$\vec{x} : \vec{A} \vdash N : B$}
    \RightLabel{${\bf let}_{\textbf{K}}$}
    \BinaryInfC{$\Gamma \vdash {\bf let \:\: pure} \:\: \vec{N} = \vec{M} \:\: {\bf in} \:\: N : {\bf K} B$}
  \end{prooftree}
    \end{flushright}
  \end{minipage}
  \end{defin}

@@

@@

  2) $(M N) [x := P] = M[x := P] N [x := P]$;

@@

  In $\lambda \textbf{K}$ we have the following computational rules:

  \begin{defin} $\beta$-reduction rules for \emph{$\lambda \textbf{K}$}.

$\begin{array}{lll}
&\text{1) } (\lambda x. M) N \rightarrow_{\beta} M [x := N]& \\
&\text{2) } \pi_1 \langle M, N \rangle \rightarrow_{\beta} M& \\
&\text{3) } \pi_2 \langle M, N \rangle \rightarrow_{\beta} N& \\
&\text{4) } {\bf let \: pure \: } \vec{y} = \vec{M_2} {\bf \: in \: } ({\bf let \: pure \: } \vec{x} = \vec{M_1} {\bf \: in \: }  N) \rightarrow_{\beta}& \\
&\quad {\bf let \: pure \: } \vec{x} = ({\bf let \: pure \: } \vec{y} = \vec{M_2} \: {\bf in} \: \vec{M_1}) \: {\bf in} \: {\bf let \: pure \: } \vec{x} = \vec{x} {\bf \: in \:} N&
\end{array}$

@@

  Induction on derivation of $\Gamma \vdash M : \alpha$. Let us assume $\Gamma \subseteq \Delta$.

  1) Let $\Gamma \vdash x : A$, such that $\Gamma = \Delta, x : A$ and $\Theta \subseteq \Gamma$.
  Let $\Sigma = \Theta \setminus \Gamma$, or, which is the same, $\Sigma = \Theta \setminus \Delta, x :
  A$, then $\Sigma, \Delta, x : A \vdash x : \alpha$, or, $\Theta \vdash x : A$.

@@

  If $\Gamma \vdash {\bf pure \: } M : \textbf{K}A$, then $\Gamma \vdash M : A$ by generation and, by
  hypothesis, $\Theta \vdash M : A$, so $\Theta \vdash {\bf pure \: } M : \textbf{K}A$ by applying
  $\textbf{K}_I$-rule.

  3) Let $\Gamma \vdash {\bf let \: pure \:} \vec{x} = \vec{M} {\bf \: in \:} N : \textbf{K}B$ and $\Gamma \subseteq \Delta$.

  By generation $\Gamma \vdash \vec{M} : \textbf{K}\vec{A}$ and $\vec{x} : \vec{A} \vdash N : \textbf{K}B$.

  By hypothesis we have $\Delta \vdash \vec{M} : \textbf{K}\vec{A}$. So $\Delta \vdash {\bf let \: pure \:} \vec{x} = \vec{M} {\bf \: in \:} N : \textbf{K}B$ by ${\bf let}_{\textbf{K}}$.

@@

  Induction by derivation. We consider the base of induction and the case with  ${\bf let}_{\textbf{K}}$. The rest
  cases are
  proven by the same way.

@@

  1) Let $\Gamma \vdash x : A$, where $\Gamma = \Delta, x : A$, $x \in \mathbb{V}$.

@@

  2) Let $\Gamma \vdash {\bf let \: pure \:} \vec{x} = \vec{M} {\bf \: in \:} N : \textbf{K}B$.

  By generation $\Gamma \vdash \vec{M} : \textbf{K}\vec{A}$ and $\vec{x} : \vec{A} \vdash N : \textbf{K}B$.

  By hypothesis $\Gamma \uparrow FV(\vec{M}) \vdash \vec{M} : \textbf{K}\vec{A}$.

  So $\Gamma \uparrow FV(\vec{M}) \vdash {\bf let \: pure \:} \vec{x} = \vec{M} {\bf \: in \:} N : \textbf{K}B$ by ${\bf let}_{\textbf{K}}$.

@@

  If $\Gamma, x : A \vdash M : B$  and $\Gamma \vdash N : A$, then $\Gamma \vdash (M [x := N]) :
  B $

@@

  If $\Gamma, x : \alpha \vdash {\bf pure \: } M : \textbf{K}\beta$.

@@

  So, by induction hypothesis, $\Gamma \vdash (M [x := N]) : B $.
  Then $\Gamma \vdash {\bf pure \: } (M [x := N]) : \textbf{K}B$ by $\textbf{K}_I$, but ${\bf pure \: } (M [x := N]) =
  ({\bf pure \: } M [x := N])$ by substitution definition.

  So $\Gamma \vdash ({\bf pure \: } M [x := N]) : \textbf{K}B$

  2) Let $\Gamma, y : C \vdash {\bf let \: pure \:} \vec{x} = \vec{M} {\bf \: in \:} N : \textbf{K}B$ and $\Gamma \vdash P : C$.

  By generation $\Gamma, y : C \vdash \vec{M} : \textbf{K}\vec{A}$ and $\vec{x} : \vec{A} \vdash N : \textbf{K}B$.

  By assumption $\Gamma \vdash \overrightarrow{M [y := P]} : \textbf{K}A$.

  So $\Gamma \vdash {\bf let \: pure \:} \vec{x} = \overrightarrow{M [y := P]} {\bf \: in \:} N : \textbf{K}B$ by ${\bf let}_{\textbf{K}}$.

@@

  \begin{defin} Type substitution

    The substitution of type $C$ for type variable $B$ in some type $A$ is defined by:

    1) $A [ A := B ] = B$;

    2) $A^{'} [A := C] = A^{'}$;

    3) $(A_1 \alpha A_2) [B := C] = A_1 [B := C] \alpha A_2 [A := B]$, where $\alpha \{ \times, \to \}$;

    4) $(\textbf{K}A) [B := C] = \textbf{K}(A [B := C])$.

  \begin{lemma} Substitution for types.

    Let $\Gamma \vdash M : A$, then $\Gamma [B := C] \vdash M : A [B := C]$, where $\Gamma [B := C]$ denotes $\{ x_1 : A_1 [B := C],\dots, x_n : A_n [B := C] \}$.

  1) Let $\Gamma, x : A \vdash x : A$, so $\Gamma [B := C], x : A [B := C] \vdash x : A [B := C]$.

  2) Let $\Gamma \vdash {\bf pure \: } M : \textbf{K}A$. By generation $\Gamma \vdash M : A$.

  By assumption $\Gamma [B := C] \vdash M : A [B := C]$.

  By $\textbf{K}_I$, $\Gamma [B := C] \vdash {\bf pure \:} M : \textbf{K} (A [B := C])$.

  But, by substitution definition, $\textbf{K} (A [B := C]) = (\textbf{K}A) [B := C]$.

  Then $\Gamma [B := C] \vdash {\bf pure \:} M : (\textbf{K}A) [B := C]$.

  3) $\Gamma \vdash {\bf let \: pure} \: \vec{x} = \vec{M} \: {\bf in} \: M : {\bf K} B$.

  By generation $\Gamma \vdash \vec{M} : \textbf{K}\vec{A}$ and $\vec{x} : \vec{A} \vdash N : \textbf{K}B$.

  $\Gamma [D := C] \vdash \vec{M} : \textbf{K}\overrightarrow{A [D := C]}$ and $\vec{x} : \overrightarrow{A [D := C]} \vdash M : B [D := C]$ by assumption.

  Then $\Gamma [D := C] \vdash {\bf let \: pure } \: \vec{x} = \vec{M} \: {\bf in} \: M : {\bf } (B [D := C])$.

@@

@@

@@

  \bibitem{Baren2} H.P. Barendregt. The Lambda Calculus --- Its Syntax and Semantics. Studies in Logic and
  the Foundations of Mathematics, vol. 103. Amsterdam: North-Holland, 1985.
  \end{thebibliography}

>>>>

\begin{lemma}
  $ $
  If $\text{IEL}^{-} \vdash A$, then $\text{NIEL}^{-} \vdash A$.
\end{lemma}

\begin{proof}
  Straightforward derivation of modal axioms in $\text{NIEL}^{-}$. We consider this derivation below using terms.
\end{proof}

@@

  \begin{center}
  \begin{prooftree}
  \AxiomC{$ $}
  \RightLabel{\scriptsize{ax}}
  \UnaryInfC{$\Gamma , x : \alpha \vdash x : \alpha$}
  \end{prooftree}
  \end{center}

  \begin{minipage}{0.5\textwidth}
    \begin{flushleft}
  \begin{prooftree}
  \AxiomC{$\Gamma, x : \alpha \vdash M : \beta$}
  \RightLabel{$\rightarrow_i$}
  \UnaryInfC{$\Gamma \vdash \lambda x. M : \alpha \to \beta$}
  \end{prooftree}

  \begin{prooftree}
  \AxiomC{ $\Gamma \vdash x : \alpha$ }
  \AxiomC{ $\Gamma \vdash y : \beta$ }
  \RightLabel{$\times_i$}
  \BinaryInfC{$\Gamma \vdash (x, y) : \alpha \times \beta$}
  \end{prooftree}

  \begin{prooftree}
    \AxiomC{$\Gamma \vdash x : \alpha$}
    \RightLabel{$\textbf{K}_I$}
    \UnaryInfC{$\Gamma \vdash {\bf pure \: } \: x : \textbf{K}\alpha $}
  \end{prooftree}
    \end{flushleft}
  \end{minipage}

  \begin{minipage}{0.5\textwidth}
    \begin{flushright}
  \begin{prooftree}
  \AxiomC{$\Gamma \vdash f : \alpha \to \beta$}
  \AxiomC{$\Gamma \vdash x : \alpha$}
  \RightLabel{$\rightarrow_e$}
  \BinaryInfC{$\Gamma \vdash fx : \beta$}
  \end{prooftree}

  \begin{prooftree}
  \AxiomC{ $\Gamma \vdash p : \alpha_1 \times \alpha_2$ }
  \RightLabel{$\times_e$, $i \in \{ 1, 2 \}$}
  \UnaryInfC{$\Gamma \vdash \pi_i p : \alpha_i$}
  \end{prooftree}

  \begin{prooftree}
    \AxiomC{$\Gamma \vdash \vec{M} : {\bf K} \vec{A}$}
    \AxiomC{$\vec{x} : \vec{A} \vdash M : B$}
    \RightLabel{$\text{let}_{\bf K}$}
    \BinaryInfC{$\Gamma \vdash {\bf let \: pure \:} \vec{x} = \vec{M} {\: \bf in \: } M : {\bf K} B$}
  \end{prooftree}
    \end{flushright}
  \end{minipage}
  \end{defin}

@@

  $\text{let}_{\bf K}$ is similar to $\Box_I$-rule in typed lambda calculus for intuitionistic normal modal logic {\bf IK}, which is described in \cite{ModalK}.

@@

@@

  \begin{defin} $\beta$-reduction and $\eta$-reduction rules for \emph{$\lambda \textbf{K}$}.

  1) $(\lambda x. M) N \rightarrow_{\beta} M [x := N]$;

  2) $\pi_1 \langle M, N \rangle \rightarrow_{\beta} M$;

  3) $\pi_2 \langle M, N \rangle \rightarrow_{\beta} N$;

  4) $\begin{array}{llll}
  & {\bf let \: pure \:} \langle \vec{x}, y, \vec{z} \rangle = \langle \vec{M}, {\bf let \: pure \: } \vec{w} = \vec{N} {\bf \: in \: } Q, \vec{P} \rangle {\: in \:} R \rightarrow_{\beta} & \\
  & {\bf let \: pure \:} \langle \vec{x}, \vec{w}, \vec{z} \rangle = \langle \vec{M}, \vec{N}, \vec{P} \rangle \: { \bf in } \: R [y := Q]
  \end{array}$

  5) ${\bf pure \:} ((\lambda x. M) N) \rightarrow_{\beta} {\bf pure \: } (M [x := N])$;

  6) ${\bf pure \:} (\pi_i \langle M_1, M_2 \rangle) \rightarrow_{\beta} {\bf pure \:} M_i$, where $i \in \{ 1, 2 \}$.

  7) $\begin{array}{llll}
  & {\bf pure \:} ({\bf let \: pure \:} \langle \vec{x}, y, \vec{z} \rangle = \langle \vec{M}, {\bf let \: pure \: } \vec{w} = \vec{N} {\bf \: in \: } Q, \vec{P} \rangle {\: in \:} R) \rightarrow_{\beta} & \\
  & {\bf pure \:} ({\bf let \: pure \:} \langle \vec{x}, \vec{w}, \vec{z} \rangle = \langle \vec{M}, \vec{N}, \vec{P} \rangle \: { \bf in } \: R [y := Q])
  \end{array}$

  8) $\lambda x. f x \rightarrow_{\eta} f$;

  9) $\langle \pi_1 P, \pi_2 P \rangle \rightarrow_{\eta} P$;

  10) ${\bf let \: pure \:} x = M {\: \bf in \: } x \rightarrow_{\eta} M$;

  11) ${\bf pure \:} (\lambda x. f x) \rightarrow_{\eta} {\bf pure \:} f$;

  12) ${\bf pure \:} (\langle \pi_1 P, \pi_2 P \rangle) \rightarrow_{\eta} {\bf pure \:} P$;

  13) ${\bf pure \:} ({\bf let \: pure \:} x = M {\: \bf in \: } x) \rightarrow_{\eta} {\bf pure \:} M$.
  \end{defin}

@@

  Induction on derivation of $\Gamma \vdash M : A$. Let us assume $\Gamma \subseteq \Delta$.

  1) Let $\Gamma \vdash x : A$, such that $\Gamma = \Delta, x : A$ and $\Theta \subseteq \Gamma$.
  Let $\Sigma = \Theta \setminus \Gamma$, or, which is the same, $\Sigma = \Theta \setminus \Delta, x :
  A$, then $\Sigma, \Delta, x : A \vdash x : A$, or, $\Theta \vdash x : A$.

@@

  By generation $\Gamma \vdash M : A$

  By hypothesis, $\Theta \vdash M : A$, so $\Theta \vdash {\bf pure \: } M : \textbf{K}A$ by applying
  $\textbf{K}_I$-rule.

  3) Let $\Gamma \vdash {\bf let \: pure \:} \vec{x} = \vec{M} {\: \bf in \:} N : {\bf K} B$ and $\Gamma \subseteq \Theta$.

  By generation $\Gamma \vdash \vec{M} : {\bf K} \vec{A}$ and $\vec{x} : \vec{A} \vdash N : B$.

  By assumption $\Theta \vdash \vec{M} : {\bf K} \vec{A}$.

  Hence $\Theta \vdash {\bf let \: pure \:} \vec{x} = \vec{M} {\: \bf in \:} N : {\bf K} B$.

@@

@@

  1) Let $\Gamma \vdash x : A$, where $\Gamma = \Delta, x : \alpha$, $x \in \mathbb{V}$.

@@

  2) Let $\Gamma \vdash {\bf pure \: } M : {\bf K} A$.

  By generation, $\Gamma \vdash M : A$ and $\Gamma \uparrow FV(M) \vdash M : A$ by hypothesis.

  So $\Gamma \uparrow FV(M) \vdash {\bf pure \: } M : {\bf K} A$ by ${\bf K}_I$.

  3) Let $\Gamma \vdash {\bf let \: pure \:} \vec{x} = \vec{M} {\: \bf in \:} N : {\bf K} B$.

By generation, $\Gamma \vdash \vec{M} : {\bf K} \vec{A}$ and $\vec{x} : \vec{A} \vdash N : B$.

By assumption, $\Gamma \uparrow FV(\vec{M}) \vdash \vec{M} : {\bf } \vec{A}$.

By $\text{let}_{\bf K}$, $\Gamma \uparrow FV(\vec{M}) \vdash {\bf let \: pure \:} \vec{x} = \vec{M} {\: \bf in \:} N : {\bf K} B$

@@

  If $\Gamma, x : A \vdash M : B$  and $\Gamma \vdash N : A$, then $\Gamma \vdash (M [x := N]) : B$

@@

@@

  By assumption, $\Gamma \vdash (M [x := N]) : B$

  Then $\Gamma \vdash {\bf pure \: } (M [x := N]) : \textbf{K}B$ by $\textbf{K}_I$.

  But ${\bf pure \: } (M [x := N]) = ({\bf pure \: } M [x := N]) $ by substitution definition, so
  $\Gamma \vdash ({\bf pure \: } M [x := N]) : \textbf{K}B$

  2) Let $\Gamma, y : A \vdash {\bf let \: pure \:} \vec{x} = \vec{M} {\: \bf in \:} N : {\bf K} B$ and $\Gamma \vdash N : A$.

  By generation, $\Gamma, y : A \vdash \vec{M} : {\bf K} \vec{A}$ and $\vec{x} : \vec{A} \vdash N : B$.

  By hypothesis, $\Gamma \vdash \vec{M}[x := N] : {\bf K} \vec{A}$.

  Hence $\Gamma \vdash {\bf let \: pure \:} \vec{x} = \vec{M}[x := N] {\: \bf in \:} N : {\bf K} B$.

@@

@@

We consider only modal $\beta$-reduction rules. The general statement for $\twoheadrightarrow_{\beta}$ follows from transitivity of multi-step $\beta$-reduction.

@@

1) Let $\Gamma \vdash {\bf let \: pure \:} \langle \vec{x}, y, \vec{z} \rangle = \langle \vec{M}, {\bf let \: pure \: } \vec{w} = \vec{N} {\bf \: in \: } Q, \vec{P} \rangle {\: in \:} R : {\bf K }B$

By generation we have $\Gamma \vdash \vec{M} : {\bf K}\vec{A_1}$, $\Gamma \vdash {\bf let \: pure \: } \vec{w} = \vec{N} {\bf \: in \: } Q : {\bf K}\vec{A_2}$, $\Gamma \vdash \vec{P} : {\bf K}\vec{A_3}$
and $\vec{x} : \vec{A_1}, y : A_2, \vec{z} : \vec{A_3} \vdash R : B$.

If $\Gamma \vdash {\bf let \: pure \: } \vec{w} = \vec{N} {\bf \: in \: } Q : {\bf K}\vec{A_2}$, then

$\Gamma \vdash \vec{N} : {\bf K}\vec{A_4}$ and $\vec{w} : \vec{A_4} \vdash Q : A_2$.

Then $\vec{x} : \vec{A_1}, \vec{w} : \vec{A_4}, \vec{z} : \vec{A_3} \vdash R [y := Q] : B$ by substitution lemma and weakening.

Hence $\Gamma \vdash {\bf let \: pure \:} \langle \vec{x}, \vec{w}, \vec{z} \rangle = \langle \vec{M}, \vec{N}, \vec{P} \rangle \: { \bf in } \: R [y := Q] : {\bf K} B$
by $let_{{\bf K}}$.

2) Let $\Gamma \vdash {\bf pure } ((\lambda x. M) N) : {\bf K} B$.

By generation $\Gamma \vdash (\lambda x. M) N : B$, but $\Gamma \vdash M [x := N] : B$, then, by ${\bf K}_I$, $\Gamma \vdash {\bf pure \:} (M [x := N]) : {\bf K} B$.

3) Let $\Gamma \vdash {\bf pure } (\pi_i \langle M_1, M_2 \rangle) : {\bf K} A_i$, where $i \in \{ 1,2 \}$.

By generation $\Gamma \vdash \pi_i \langle M_1, M_2 \rangle : A_i$ and $\Gamma \vdash M_i : A_i$.

Hence $\Gamma \vdash {\bf pure \: } M_i : { \bf K } A_i$ by ${\bf K}_I$
  \end{proof}

@@

  \bibitem{Baren2} H.P. Barendregt. The Lambda Calculus --- Its Syntax and Semantics. Studies in Logic and
  the Foundations of Mathematics, vol. 103. Amsterdam: North-Holland, 1985.

  \bibitem{ModalK} Yoshihiko KAKUTANI, A Curry-Howard Correspondence for Intuitionistic Normal Modal Logic, Computer Software, Released February 29, 2008, Online ISSN , Print ISSN 0289-6540.
  \end{thebibliography}
